---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Introduction
The data is from a personal activity monitoring device that collects data at 5 minute intervals through out the day. The data consists of two months of data during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. Activity monitoring data [52K] The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data
```{r loaddata}
unzip(zipfile="activity.zip")
data <- read.csv("activity.csv")
data$date  <- as.Date.character(data$date)

```

## Mean and median number of steps taken each day
### What is mean total number of steps taken per day?

##### 1. Calculate the total number of steps taken per day (ignore the missing values in the dataset.)

```{r totalStepsByDay}

library(ggplot2)
dailySteps <- aggregate(x=list(steps=data$steps), by=list(date=data$date), FUN=sum, na.rm=TRUE)

ggplot(data=dailySteps, aes(x=date, y=steps)) +  geom_point() +
  geom_line() +
  xlab("date") +
  ylab("total number of steps per day")
```

##### 2. Make a histogram of the total number of steps taken each day
```{r }
steps_per_day <- tapply(data$steps, data$date, FUN=sum, na.rm=TRUE)
qplot(steps_per_day, binwidth=500, xlab="Daily histogram of the total number of steps taken for 61 days")

dailyMean <- mean(dailySteps$steps, na.rm=TRUE)
dailyMedian <- median(dailySteps$steps, na.rm=TRUE)

```

##### 3. Mean and median number of steps taken each day:
- The average daily steps is `r dailyMean`, the median daily steps is `r dailyMedian`.

## What is the average daily activity pattern?
##### 1. Make a time series plot of the average number of steps taken
```{r }
dailyAverages <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval), FUN=mean, na.rm=TRUE)
dailyAverages$intervalTime  <- gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", dailyAverages$interval)
ggplot(data=dailyAverages, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5-minute interval") +
  ylab("average number of steps taken across all 61 days")
```  

##### 2. On average which 5-minute interval contains the maximum number of steps?
```{r }
maxStep <- dailyAverages[which.max(dailyAverages$steps),]

```
The max step `r round(maxStep$steps)` happened at `r maxStep$intervalTime` of the day

## Imputing missing values
Code to describe and show a strategy for imputing missing data

##### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
missingSteps <- length(which(is.na(data$steps)))
```

- There are `r missingSteps` number of missing values.

##### 2. Use impute function in the Hmisc package to fill in NAs with the mean value of its 5-minute interval

##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, results='hide', warning=FALSE, message=FALSE}
library(Hmisc)
imputedData <- data
imputedData$steps <- impute(data$steps, fun=mean)
```

##### 4. Plot histogram of the total number of steps taken each day after missing values are imputed
```{r}
stepsImputedDaily <- tapply(imputedData$steps, imputedData$date, sum)
qplot(stepsImputedDaily, xlab='Daily histogram of the total steps (Imputed)', ylab='Frequency', binwidth=500)

meanImputedStep <- as.character(round(mean(stepsImputedDaily)))
medianImputedStep <- as.character(round(median(stepsImputedDaily)))
```

Mean: `r meanImputedStep` and Median:  `r medianImputedStep` of the imputed data

## Are there differences in activity patterns between weekdays and weekends?
Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

##### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
  
```{r}
imputedData$dayType <-  ifelse(as.POSIXlt(imputedData$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

##### 2. Make a panel plot containing a time series plot

```{r}
averagedimputedData <- aggregate(steps ~ interval + dayType, data=imputedData, mean)
ggplot(averagedimputedData, aes(interval, steps)) + 
  geom_line() + 
  facet_grid(dayType ~ .) +
  xlab("5-minute interval") + 
  ylab("avarage number of steps")
```
